---
title: "Impact of Maternal Smoking on Infant Birthweight"
author: "Author 1 and Author 2"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Header

## Author Contributions

Author 1:

Author 2:

## Use of GPT

ChatGPT was used as a substitute for documentation for R. Since we were unfamiliar with R, we asked ChatGPT how to use R in certain methods in order to find and filter out conditions in the dataset. We additionally used GPT to analyze reasoning and to confirm what we thought was correct about the dataset, as well as to identify extra questions that could be answered for our advanced analysis.

# Introduction

## Main Research Questions

1.  

## Outline

The remainder of the report will go through a basic analysis of the data, including our cleaning methods, basic analysis on various variables in the study, and more. Additionally, we will analysis the questions posed above, along with the conclusions that we came up with in our data. We will also pose an advanced analysis question based on the relationship between the gestation period of a pregnancy and see how the relationship between a child's birthweight and their mother's smoking-status are intertwined.

# Basic Analysis

## Data Processing and Summaries

### Methods

```{r data-processing}
# Your R code for data processing
```

### Analysis

```{r data-summary}
# Your R code for data summary
```

### Conclusions

Your conclusions about data processing and summaries.

## Question 1

### Methods

```{r q1-methods}
# Your R code for methods related to Question 1
```

### Analysis

```{r q1-analysis}
# Your R code for analysis related to Question 1
```

### Conclusions

Your conclusions for Question 1.

## Question 2

### Methods

```{r q2-methods}
# Your R code for methods related to Question 2
```

### Analysis

```{r q2-analysis}
# Your R code for analysis related to Question 2
```

### Conclusions

Your conclusions for Question 2.

## Question 3

### Methods

```{r q3-methods}
# Your R code for methods related to Question 3
```

### Analysis

```{r q3-analysis}
# Your R code for analysis related to Question 3
```

### Conclusions

Your conclusions for Question 3.

## Question 4

### Methods

```{r q4-methods}
# Your R code for methods related to Question 4
```

### Analysis

```{r q4-analysis}
# Your R code for analysis related to Question 4
```

### Conclusions

Your conclusions for Question 4.

# Advanced Analysis

## Additional Research Question

### Methods

```{r advanced-methods}
# Your R code for methods related to the advanced question
```

### Analysis

```{r advanced-analysis}
# Your R code for analysis related to the advanced question
```

### Conclusions

Your conclusions for the advanced analysis.

# Conclusions and Discussion

## Summary of Findings

Reprise the questions and goals of the analysis stated in the introduction. Summarize the findings and compare them to the original goals.

## Discussion

Additional observations or details gleaned from the analysis section. Discuss relevance to the science and other studies, if applicable. Address data limitations. Raise new questions and suggest future work.

# Appendix

Include any additional technical details, tables, or figures that support your analysis but would disrupt the flow of the main text if included earlier.

```{r appendix-code, eval=FALSE}
# Additional R code or output can be included here
```
