---
title: "Finding the Relationship between Density and Gain"
author: "Author 1 and Author 2"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Header

## Author Contributions

Author 1: Contributed to questions 1, 4, 5, and did the formatting for the pdf

Author 2: Contributed to questions 2, 3, and 6, along with doing the advanced analysis.

## Use of GPT

ChatGPT was used as a substitute for documentation for R. Since we were unfamiliar with R, we asked ChatGPT how to use R in certain methods in order to find and filter out conditions in the dataset. We additionally used GPT to analyze reasoning and to confirm what we thought was correct about the dataset, as well as to identify extra questions that could be answered for our advanced analysis.

# Introduction

The data used in this analysis is from the 2020 Kaggle "Most Popular & Widely Used Machine Learning" Survey. It details questions about Kaggle's users, as well as their most used machine learning languages, platforms, education levels, and more. Our analysis today focuses on identifying key relationships within the data and to notice if there are any interesting patterns that we can find within it. 

## Main Research Questions

1) Analyze the relationship between age groups and highest education levels amongst the survey respondents. How does educational attainment vary across those in different age brackets. (can use chi-square test)
2) Investigate the correlation between years of programming experience and programming languages used, if any. Do more experienced users tend to use a wider variety of programming languages, or specific programming languages. (can use the release date of certain languages as a spearhead)
3) Examine the global distribution of survey participants and their professional jobs, extrapolating this to the general population of the world. Are certain jobs more popular in certain countries?
4) Examine differences in companies that utilize more machine learning in their operations. Do companies that use machine learning heavily tend to be larger or more successful? (can use number of employees and salaries as a proxy for company success)

## Question 1) Age-Education Relationship

One thing we want to test in regards to the age-education relationship in this graph is that we want to compare it to the education attainment for adults 25 and older in the United States. This has the following distribution: 

  * 30.5% have no formal education
  * 26.8% have some college/university experience
  * 25.8% have a bachelor's degree
  * 10.9% have a master's degree
  * 3.7% have a professional degree
  * 2.3% have a doctoral degree

### Methods

```{r, echo = FALSE}
library(ggplot2)

data <- read.csv("data.csv", header = TRUE, stringsAsFactors = FALSE)
cleaned_data <- data[!(data$Q4 %in% c("", "I prefer not to answer")), ]
ages <- cleaned_data$Q1[2:nrow(cleaned_data)]
education <- cleaned_data$Q4[2:nrow(cleaned_data)]

# create actual education data to compare towards
education_data <- data.frame(
  education_level = c(
    "Doctoral degree",
    "Professional degree",
    "Master’s degree",
    "Bachelor’s degree",
    "Some college/university study without earning a bachelor’s degree",
    "No formal education past high school"
  ),
  percentage = c(2.3, 3.7, 10.9, 25.8, 26.8, 30.5)
)

# identify the order of education
education_levels <- c(
  "Doctoral degree",
  "Professional degree",
  "Master’s degree",
  "Bachelor’s degree",
  "Some college/university study without earning a bachelor’s degree",
  "No formal education past high school"
)
education <- factor(education, levels = education_levels, ordered = TRUE)

# creates contingency table for ages & education
contingency_table <- table(ages, education)
age_edu_table <- as.data.frame(contingency_table)
colnames(age_edu_table) <- c("AgeGroup", "EducationLevel", "Count")

# Create a stacked chart
ggplot(age_edu_table, aes(x = AgeGroup, y = Count, fill = EducationLevel)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Educational Attainment Across Age Groups",
       x = "Age Group",
       y = "Count",
       fill = "Education Level") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Calculate the total observed counts for each education level
observed_counts <- colSums(contingency_table)

# Calculate the expected proportions from education_data
expected_proportions <- education_data$percentage / 100

# Calculate the expected counts based on the total sample size
total_sample_size <- sum(observed_counts)
expected_counts <- expected_proportions * total_sample_size

# Perform Chi-square Goodness of Fit test
chi_square_test <- chisq.test(x = observed_counts, 
                               p = expected_proportions)

# Print the results
print(chi_square_test)
```  

### Analysis

Looking through this graph, we can notice a lot of things. For instance, it can be noticed that there is an obvious trend of younger users who took this survey, with the peak of the users falling from the 18-21 range to the 25-29 range. This slowly falls off, with each age range slowly losing ground on the majority of users. Other patterns to notice unclude the influx of people with a bachelor's degree, with the visible majority being placed at the age range of 18-21, and slowly lowering down percent wise as age increases. On the other hand, degrees like masters, professional, and doctoral are very low in the 18-21 age range, which makes sense intuitively, because these degrees typically require the previous ones to acquire. This means that those in the 18-21 and 22-24 age ranges very rarely have these advanced degrees, and the age range of 25-29 is the first age range we actually see a significant amount of these education levels. Otherwise, other education levels to note include the fact that master's degrees are found often at the 22-24 age range and beyond, comprising the most of all the education levels after this age. Additionally, when comparing the given probabilities to the true distribution of education levels in the United States for ages 25 and above, we can notice that the Chi-Square Test has a p-value of 2.2e-16, which means that our data is significantly different than the actual distribution of education levels amongst adults in the United States. 

### Conclusion

Further analysis of the data and the graph created above shows that the educational attainment for those who took this survey, and by connotation, are interested in Machine Learning, is very different than the actual educational attainment in the United States. This somewhat makes sense; machine learning is a typically much more education-based topic that people have to learn, so those who want to learn it and apply it in their lives typically go to college and even higher education to achieve their goals. The other things we noticed also make sense; younger people rarely have degrees above a bachelor's, because they don't have the necessary amount of time to get those degrees. Additionally, most of the ages in the graph tend to be younger, skewing right, which makes sense, because machine learning has only become an especially popular topic recently, which maybe be reflected in our data. 