# DSC 190: Chapter 2

#-------------------------------------------------------------------------------
# Case study 1

# Load data
```{r}
dat = read.table('babies.txt', header=T)
head(dat)
```

# Histogram of mothers' heights
```{r}
hist(dat$height)
hist(dat$height, breaks = 20)
hist(dat$height, breaks = 50)
hist(dat$height, breaks = 100)
```

# Remove missing data
```{r}
ind = which(dat$height > 90)
dat[ind,]
dat.valid = dat[-ind,]
hist(dat.valid$height, breaks = 20)
```
#-------------------------------------------------------------------------------
# The normal curve

# Standard normal
```{r}
curve(dnorm(x), -4, 4)
abline(h = 0)
abline(v = -1, col='blue'); pnorm(-1)
abline(v = 1, col='blue'); pnorm(1)
pnorm(1) - pnorm(-1)
abline(v = -2, col='red'); pnorm(-2)
abline(v = 2, col='red'); pnorm(2)
pnorm(2) - pnorm(-2)
```

# Other parameters
```{r}
mu = 5; sigma = 2
n = 1000
X = rnorm(n, mean = mu, sd = sigma)
hist(X, breaks = 50, probability=T)
curve(dnorm(x), -4, 4, col='red', lwd=2, add=T)
curve(dnorm(x, mean = mu, sd = sigma), -1, 13, col='blue', lwd=2, add=T)
Xbar = mean(X)
Xbar
S = sd(X)
S
Z = (X - Xbar)/S
hist(Z, breaks = 50, probability=T)
curve(dnorm(x), -4, 4, col='red', lwd=2, add=T)
```

# CHDS data
```{r}
Z = (dat.valid$height - mean(dat.valid$height)) / sd(dat.valid$height)
F = ecdf(Z)
F(3) - F(-3)
F(2) - F(-2)
F(1) - F(-1)
```
#-------------------------------------------------------------------------------
# Skewness
skew = function(X) {
  Xbar = mean(X); S = sd(X)
  Z = (X - Xbar)/S
  mean(Z^3)
}

# Simulate
mu = 5; sigma = 2
n = nrow(dat)
X = rnorm(n, mean = mu, sd = sigma)
Z = (X - mean(X)) / sd(X)
skew(X)
skew(Z)

# Kurtosis
kurt = function(X) {
  Xbar = mean(X); S = sd(X)
  Z = (X - Xbar)/S
  mean(Z^4)
}

# Simulate
mu = 5; sigma = 2
n = nrow(dat)
X = rnorm(n, mean = mu, sd = sigma)
Z = (X - mean(X)) / sd(X)
kurt(X)
kurt(Z)

# Repeating the above simulation N=1000 times will give a distribution of possible skewness and kurtosis values.
# Skewness and Kurtosis are examples. The same technique can be applied to any statistic.

kurt(dat.valid$height)

#-------------------------------------------------------------------------------
# QQ plot
X = rnorm(n, mean = mu, sd = sigma)
Z = (X - mean(X)) / sd(X)
qqnorm(Z)
abline(0, 1, col='red', lwd=2)
qqnorm(X)
abline(0, 1, col='red', lwd=2)
abline(mu, sigma, col='blue', lwd=2)

# Example: uniform
U = runif(1000, min = -1, max = 1)
hist(U, breaks = 50, probability = T, xlim = c(-3,3))
curve(dnorm(x), -4, 4, col='red', lwd=2, add=T)
qqnorm(U)
abline(0, 1, col='red', lwd=2)

# Example: exponential
W = rexp(1000, rate = 1)
hist(W, breaks = 50, probability = T, xlim = c(-3,3))
curve(dnorm(x), -4, 4, col='red', lwd=2, add=T)
qqnorm(W)
abline(0, 1, col='red', lwd=2)

