```{r, echo = FALSE}
data <- read.table("hcmv.txt", header=TRUE)
```

## Question 1: Compare Data with Random Scattering of Palindromes

The dataset contains the locations of palindromic sequences in a series of 229,354 base pairs. We will do multiple simulations of random locations of palindromic sites along these 229,354 base pairs and compare it to the real data.

```{r, echo = FALSE}
set.seed(236)

generate_palindrome_sites <- function() {
  sort(sample(1:229354, 296, replace=FALSE))
}

simulated_palindrome_data <- replicate(1000, generate_palindrome_sites())

calc_summary_stats <- function(locations) {
  list(
    mean = mean(locations),
    median = median(locations),
    variance = var(locations),
    std_dev = sd(locations)
  )
}

simulated_stats <- apply(simulated_palindrome_data, 2, calc_summary_stats)
real_stats <- calc_summary_stats(data$location)

mean_simulated_means <- mean(sapply(simulated_stats, function(x) x$mean))
mean_simulated_medians <- mean(sapply(simulated_stats, function(x) x$median))
mean_simulated_variances <- mean(sapply(simulated_stats, function(x) x$variance))
mean_simulated_std_devs <- mean(sapply(simulated_stats, function(x) x$std_dev))

comparison_table <- data.frame(
  Statistic = c("Mean", "Median", "Variance", "Standard Deviation"),
  Real_Locations = c(real_stats$mean, real_stats$median, real_stats$variance, real_stats$std_dev),
  Simulated_Locations = c(mean_simulated_means, mean_simulated_medians, mean_simulated_variances, mean_simulated_std_devs)
)

print(comparison_table)
```

To qualitatively compare the real palindrome locations to the simulated locations, I plot a histogram of the means of each simulated palindrome sequence and analyze where the real mean is on the distribution. The same process is also done for the standard deviations.

```{r, echo = FALSE}
par(mfrow = c(1, 2))

simulated_means <- sapply(simulated_stats, function(x) x$mean)
hist(simulated_means, breaks = 30, col = "lightblue", main = "Simulated Mean Locations", xlab = "Mean of Palindrome Locations", cex.main = 0.8)
abline(v = real_stats$mean, col = "red", lwd = 2)
legend("topright", legend = "Real Data Mean", col = "red", lwd = 2, cex=0.5)

simulated_std_devs <- sapply(simulated_stats, function(x) x$std_dev)
hist(simulated_std_devs, breaks = 30, col = "lightgreen", main = "Simulated Standard Deviations", xlab = "Standard Deviation of Palindrome Locations", cex.main = 0.8)
abline(v = real_stats$std_dev, col = "red", lwd = 2)
legend("topright", legend = "Real Data Std Dev", col = "red", lwd = 2, cex=0.5)

par(mfrow = c(1, 1))
```

## Question 3: Compare Counts of Palindromes in Intervals to Uniform Random Scatter

```{r, echo = FALSE}
interval_size <- floor(229354 / 57)
interval_counts <- hist(data$location, breaks = seq(0, 229354, by = interval_size), plot = FALSE)$counts

interval_counts_df <- data.frame(
  Interval = 1:57,
  Count = interval_counts
)

lambda_rate <- 296 / 57
  
count_frequency_df <- as.data.frame(table(interval_counts_df$Count))
colnames(count_frequency_df) <- c("Palindrome_Count", "Intervals_Observed")
count_frequency_df$Palindrome_Count <- as.integer(as.character(count_frequency_df$Palindrome_Count))
count_frequency_df$Intervals_Expected <- dpois(count_frequency_df$Palindrome_Count, lambda_rate) * 57
count_frequency_df$Intervals_Expected <- count_frequency_df$Intervals_Expected * (57 / sum(count_frequency_df$Intervals_Expected))


par(mfrow = c(1, 2))
barplot(
  count_frequency_df$Intervals_Observed,
  names.arg = count_frequency_df$Palindrome_Count,
  col = "blue",
  xlab = "Palindrome Count",
  ylab = "Number of Intervals",
  main = "Observed Frequency of Palindrome Counts",
  cex.main = 0.8
)

barplot(
  count_frequency_df$Intervals_Expected,
  names.arg = count_frequency_df$Palindrome_Count,
  col = "red",
  xlab = "Palindrome Count",
  ylab = "Number of Intervals",
  main = "Expected Frequency of Palindrome Counts (Poisson)",
  cex.main = 0.7
)

par(mfrow = c(1, 1))
```

```{r, echo = FALSE}
chi_square_test <- chisq.test(
  x = count_frequency_df$Intervals_Observed,
  p = count_frequency_df$Intervals_Expected / sum(count_frequency_df$Intervals_Expected),
  rescale.p = TRUE
)

chi_square_test
```

Since the p-value is less than our significance level of 0.05, there is significant evidence to reject the null hypothesis, meaning the palindrome sites are not uniformly distributed along the DNA sequence according to the Poisson distribution. This implies there may be some pattern or clustering of the palindromes along the DNA.

Next, we investigate the effect of different interval sizes on the analysis and detection of uniformity.

```{r, echo = FALSE}
larger_interval_size <- floor(229354 / 10)
larger_interval_counts <- hist(data$location, breaks = seq(0, 229354, by = larger_interval_size), plot = FALSE)$counts

larger_interval_counts_df <- data.frame(
  Interval = 1:10,
  Count = larger_interval_counts
)

lambda_rate <- 296 / 10
  
larger_count_frequency_df <- as.data.frame(table(larger_interval_counts_df$Count))
colnames(larger_count_frequency_df) <- c("Palindrome_Count", "Intervals_Observed")
larger_count_frequency_df$Palindrome_Count <- as.integer(as.character(larger_count_frequency_df$Palindrome_Count))
larger_count_frequency_df$Intervals_Expected <- dpois(larger_count_frequency_df$Palindrome_Count, lambda_rate) * 10
larger_count_frequency_df$Intervals_Expected <- larger_count_frequency_df$Intervals_Expected * (10 / sum(larger_count_frequency_df$Intervals_Expected))


par(mfrow = c(1, 2), oma = c(0, 0, 2, 0))
barplot(
  larger_count_frequency_df$Intervals_Observed,
  names.arg = larger_count_frequency_df$Palindrome_Count,
  col = "blue",
  xlab = "Palindrome Count",
  ylab = "Number of Intervals",
  main = "Observed Frequency of Palindrome Counts",
  cex.main = 0.8
)

barplot(
  larger_count_frequency_df$Intervals_Expected,
  names.arg = larger_count_frequency_df$Palindrome_Count,
  col = "red",
  xlab = "Palindrome Count",
  ylab = "Number of Intervals",
  main = "Expected Frequency of Palindrome Counts (Poisson)",
  cex.main = 0.7
)

mtext("Interval Size = 22,935", outer = TRUE, cex = 1.5)

par(mfrow = c(1, 1))
```

```{r, echo = FALSE}
smaller_interval_size <- floor(229354 / 580)
smaller_interval_counts <- hist(data$location, breaks = seq(0, 229354, by = smaller_interval_size), plot = FALSE)$counts

smaller_interval_counts_df <- data.frame(
  Interval = 1:580,
  Count = smaller_interval_counts
)

lambda_rate <- 296 / 580
  
smaller_count_frequency_df <- as.data.frame(table(smaller_interval_counts_df$Count))
colnames(smaller_count_frequency_df) <- c("Palindrome_Count", "Intervals_Observed")
smaller_count_frequency_df$Palindrome_Count <- as.integer(as.character(smaller_count_frequency_df$Palindrome_Count))
smaller_count_frequency_df$Intervals_Expected <- dpois(smaller_count_frequency_df$Palindrome_Count, lambda_rate) * 580
smaller_count_frequency_df$Intervals_Expected <- smaller_count_frequency_df$Intervals_Expected * (580 / sum(smaller_count_frequency_df$Intervals_Expected))


par(mfrow = c(1, 2), oma = c(0, 0, 2, 0))
barplot(
  smaller_count_frequency_df$Intervals_Observed,
  names.arg = smaller_count_frequency_df$Palindrome_Count,
  col = "blue",
  xlab = "Palindrome Count",
  ylab = "Number of Intervals",
  main = "Observed Frequency of Palindrome Counts",
  cex.main = 0.8
)

barplot(
  smaller_count_frequency_df$Intervals_Expected,
  names.arg = smaller_count_frequency_df$Palindrome_Count,
  col = "red",
  xlab = "Palindrome Count",
  ylab = "Number of Intervals",
  main = "Expected Frequency of Palindrome Counts (Poisson)",
  cex.main = 0.7
)

mtext("Interval Size = 395", outer = TRUE, cex = 1.5)

par(mfrow = c(1, 1))
```
